<<<<<<< HEAD
---
title: "Lab 2 - Data Structures in R"
subtitle: "Environmental Data Analytics"
author: "John Fay & Luana Lima"
date: "Fall 2024"
---

## Objective
Discuss and navigate different data types in R
 - Vectors
 - Matrices
 - Arrays
 - Data frames

---

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.

-   **Matrices** 2 dimensional structure that contains elements of the same type.

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes.

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

---

## Vectors
### Creating Vectors
Add comments below that specify what kind of vectors are created. (First answer is provided.)

```{r Creating vectors}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # Vector type: Numeric
vector1
class(vector1)

vector2 <- c("one","two","three") # Vector type: Characters (factor)
vector2

vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) # Vector type: Logical
vector3
```

### Vector Operations
Add comments below to indicate what each operation does.(First answer is provided.)

```{r Vector operatons}
# Returns the 3rd item in the vector
vector1[3]

# Returns the max item in the vector
max(vector1)

# Create another vector of the values 10 thru 15
vector4 <- c(10:15)

# Make a list that combines vector 1 and 4
c(vector1, vector4)

# Multiply the couple located in each of the two lists 
vector1 * vector4

# Times 10 to each element in vector 1
vector1 * 10

# Make the characters to uppercase
toupper(vector2) 

# Negate vector 3
!vector3

```

### Exercise 1a: Vector
Find a 5-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create a vector representing the high temperature on each of the five days.
<https://www.wunderground.com/forecast/us/nc/durham/KNCDURHA138>

```{r Vector Exercise}

# Assign to the variable 'temps_F' a vector of high temperatures in F
temps_F <- c(80,82,80,75,74)

# What is the average of these temperatures 
mean(temps_F)

# Compute the difference between each temperature and the average calculated above
mean.temp <- mean(temps_F)
temps_F - mean.temp

# Subtract 32 and multiply by 5/9ths to create a vector in Celsius
temps_C <- (temps_F -32)/1.8
temps_C

```

## Matrices
###Creating Matrices
Complete the comments below:
```{r Creating Matrices}
# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # 
matrix1
#matrix1 has 5 rows and 4 columns

matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #
matrix2
#matrix2 has 5 rows and 4 columns

matrix3 <- matrix(
  data = 1:20, 
  nrow = 5, 
  ncol = 4, 
  byrow = TRUE, # return after comma continues the line
  dimnames = list(
    c("uno", "dos", "tres", "quatro", "cinco"),
    c("un", "deux", "trois", "quatre")
    )
  ) 
matrix3
```

### Matrix Operations
Add comments in the code chunk below to indicate type of vectors created and what each operation does.
```{r Matrix Operations}

matrix1[4, 1] # element in row 4, column 1

matrix1[4,  ] # elements in row 4

matrix1[ , 3] # elements in column 3

matrix1[c(12, 14)] # combined elements number 12 and 14

matrix1[c(12:14)] # combined elements from 12 to 14

matrix1[2:4, 1:3] # submatrix of rows 2 to 4, for columns 1 to 3
 
# Create a vector of 6 values
values = c(1, 26, 24, 68, 3, 44)
# Create a vector of row names
rnames <- c("R1", "R2") 
# Create a vector of column names
cnames <- c("C1", "C2", "C3") 

matrix4 <- matrix(
  data = values, 
  nrow = 2, 
  ncol = 3, 
  byrow = TRUE,
  dimnames = list(
    rnames, 
    cnames)
  ) # 

matrix4 # the element in R2 and C1

matrix4['R2','C1']
```

### Exercise 1b: Matrices
Compute the mean of column "C2" in matrix4. (Should be 14.5)
```{r Matrix Exercise}
mean(matrix4[,"C2"])
```



## Lists
### Creating Lists
```{r Creating Lists}
# Lists ---- 
# Creating Lists
list1 <- list(
  name = "Maria", 
  mynumbers = vector1, 
  mymatrix = matrix1, 
  age = 5.3)

```


### List Operations
```{r List Operations}
# List Operations

# return info of list 1
list1

# return the age part of the list
list1$age

# return the titles of the list
names(list1)

# return the first two items in teh list
list1[1:2]

# return the part of the list where name is Janelle
list1$name = "Janelle"

# edit the 4th element in the list
list1[[4]] = c(44,40)
```

## Data Frames
### Creating dataframes
```{r Creating data frames}
# Data Frames ----
# Create three vectors of equal length
d <- c(1, 2, 3, 4) # What type of vector?
e <- c("red", "white", "red", NA) # What type of vector?
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector?

# Combine the vectors into a dataframe
dataframe1 <- data.frame(d,e,f) 

# Set the column names & View
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # use view function to open the dataframe tab, MUST REMOVE BEFORE KNITTING!!!

# Add a new column to the dataframe
dataframe1$month <- c('Jan','Feb', 'Mar', 'Apr')


summary(dataframe1)
```


### Dataframe operations
```{r Dataframe Operations}
# Extract a column from the dataframe f
dataframe1$Color

# Extract a row
dataframe1[2,]

# Extract several rows
dataframe1[1:2,] 

# Extract a column
dataframe1[,1]

# Extract a column/columns from their name
out1 <- dataframe1['Color']; class(out1)

out2 <- dataframe1$Color; class(out2)

out3 <- dataframe1[c("ID","Passed")]; class(out3)